---
title: "Covid91 vaccine study Final2023"
author: "Connor Rice"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="final.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
```

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## General Methodology
Each group will be separated by filtering the data to create subsets that only contain those that fit int the category, then statistical analysis will be run on the subsets for each category. Analysis will be a two way variable analysis of the category of whether or not they received the placebo vaccine versus whether or not they were infected with Covid-19. Analysis will be done with the help of the cat-cat helper file.

$H_0$ There is no difference between the rate of infection between each of: Males, Females, LGBTQs, and Druggies who are given the placebo and those who are given the actual vaccine.
$H_a$ There is a significant difference in infection rate between each of: Males, Females, LGBTQs, and Druggies who are given the placebo versus those who are given the actual vaccine.

## Males
### Males Graphical Data
```{r include=FALSE,echo=FALSE}
Males <- FinalData %>% filter(sex=="M")

Males$treatment <- factor(Males$treatment, levels = c("placebo", "vaccine"))
```

```{r echo=FALSE}
barchartGC(~infected + treatment,data=Males)
barchartGC(~infected + treatment,data=Males, type="percent")
```

Based on the graphical data there was approximately a 50% reduction in infection rates in the general male population who received the vaccine versus those who received the placebo. However the total reduction in infections was only slight overall despite the large percentage reduction since the baseline infection rate of those who received the placebo was small to begin with.

### Males Numerical Data
```{r echo=FALSE}
table1 <- xtabs(~infected + treatment, data=Males)
rowPerc(table1)
colPerc(table1)
```

The somewhat large reduction in percent in infection rates of males between those who received the drug and those who received the placebo(1.1% of those given the actual vaccine were infected versus 2.5% of those who received the placebo) is a good initial indication that the vaccine effectively decreases overall infection rates of Covid-19 in males in concurrence with what was observed in the graphical data.

### Males Inferential Data
```{r echo=FALSE}
chisq.test(table1)
fisher.test(table1)


fish <- fisher.test(table1)


low=(fish$conf.int[1]/(fish$conf.int[1] + 1))*100
hi=(fish$conf.int[2]/(fish$conf.int[2] + 1))*100
options(digits=2)


low
hi
```

The very low p-values in the chi-squared test and fisher exact test(1x10^-13 and 5x10^-14 respectively) indicate that the difference in infection rates is indeed statistically significant which means the drug is effective at reducing infection rates in males and allows us to reject the null hypothesis. The effectiveness interval for the vaccine in males of 64%-74% is above the approximate 50% threshold that is desired for vaccine effectiveness.

### Males Conclusion and Discussion

Though the total number of infections prevented by the vaccine is relatively small it is deceptively effective, having a effectiveness interval of 64%-74% which indicates that it is a highly effective vaccine at preventing Covid-19 infections in the male population. This is confirmed by the very small p values obtained in the inferential data analysis that allowed for the rejection of the null hypothesis and the confirmation that the effectiveness interval is statistically significant.

## Females
### Females Graphical Data
```{r include=FALSE,echo=FALSE}
Females <- FinalData %>% filter(sex=="F")

Females$treatment <- factor(Females$treatment, levels = c("placebo", "vaccine"))
```

```{r echo=FALSE}
barchartGC(~infected + treatment,data=Females)
barchartGC(~infected + treatment,data=Females, type="percent")
```

Similarly to the general male data set, in the general female population data set there was approximately a 50% reduction in infection rates in the general female population who received the vaccine versus those who received the placebo. However the total reduction in infections was only slight overall despite the large percentage reduction since the baseline infection rate of those who received the placebo was small to begin with.

### Females Numerical Data
```{r echo=FALSE}
table1 <- xtabs(~infected + treatment, data=Females)
rowPerc(table1)
colPerc(table1)
```

The somewhat large reduction in percent in infection rates of females between those who received the drug and those who received the placebo(1.4% of those given the actual vaccine were infected versus 2% of those who received the placebo) is a good initial indication that the vaccine effectively decreases overall infection rates of Covid-19 in females in concurrence with what was observed in the graphical data. Though the percent reduction in infection rates is noticeably lower than the reduction seen in the male data set.

### Females Inferential Data
```{r echo=FALSE}
chisq.test(table1)
fisher.test(table1)


fish <- fisher.test(table1)


low=(fish$conf.int[1]/(fish$conf.int[1] + 1))*100
hi=(fish$conf.int[2]/(fish$conf.int[2] + 1))*100
options(digits=2)


low
hi
```

The very low p-values in the chi-squared test and fisher exact test(.0003 and .0002 respectively) indicate that the difference in infection rates is indeed statistically significant which means the drug is effective at reducing infection rates in females and allows us to reject the null hypothesis. The effectiveness interval for the vaccine in females of 55%-65% is just above the approximate 50% threshold that is desired for vaccine effectiveness.

### Females Conclusion and Discussion

Though the total number of infections prevented by the vaccine in the female population is relatively small it is deceptively effective, having a effectiveness interval of 55%-65% which indicates that it is a highly effective vaccine at preventing Covid-19 infections in the female population, though notably slightly less effective than in the male population. This is confirmed by the very small p values obtained in the inferential data analysis that allowed for the rejection of the null hypothesis and the confirmation that the effectiveness interval is statistically significant. 

## LGBTQ
### LGBTQ Graphical Data
```{r include=FALSE,echo=FALSE}
LGBTQs <- FinalData %>% filter(LGBTQ=="gay")

LGBTQs$treatment <- factor(LGBTQs$treatment, levels = c("placebo", "vaccine"))
```

```{r echo=FALSE}
barchartGC(~infected + treatment,data=LGBTQs)
barchartGC(~infected + treatment,data=LGBTQs, type="percent")
```

In the LGBTQ data set there was a noticeable difference to the two data sets of the male and female genders. The LGBTQ data set actually show an increase in Covid-19 infection rates in those who received the vaccine. However similarly to the gender based data sets the total numerical difference in infection rates is very small despite being a large percentage difference.

### LGBTQ Numerical Data
```{r echo=FALSE}
table1 <- xtabs(~infected + treatment, data=LGBTQs)
rowPerc(table1)
colPerc(table1)
```

The somewhat large increase in percent in infection rates within the LGBTQ sub-population between those who received the drug and those who received the placebo(3.5% of those given the actual vaccine were infected versus 0.66% of those who received the placebo) is a very poor initial indication that the vaccine effectively decreases overall infection rates of Covid-19 in the LGBTQ sub-population in concurrence with what was observed in the graphical data.

### LGBTQ Inferential Data
```{r echo=FALSE}
chisq.test(table1)
fisher.test(table1)


fish <- fisher.test(table1)


low=(fish$conf.int[1]/(fish$conf.int[1] + 1))*100
hi=(fish$conf.int[2]/(fish$conf.int[2] + 1))*100
options(digits=2)

low
hi
```

The very low p-values in the chi-squared test and fisher exact test(.00000007 and .00000001 respectively) indicate that the difference in infection rates is indeed statistically significant which means the drug is likely ineffective at reducing infection rates in the LGBTQ population and allows us to reject the null hypothesis. The effectiveness interval for the vaccine in the LGBTQ population of 7.6%-27% is far below the approximate 50% threshold that is desired for vaccine effectiveness.

### LGBTQ Conclusion and Discussion

The vaccine in the LGBTQ sub-population appears to be completely ineffective to the point of potentially increasing infection rates; having a effectiveness interval of 7.6%-27% which indicates that it is an effective vaccine at preventing Covid-19 infections in the LGBTQ sub-population. This is confirmed by the very small p values obtained in the inferential data analysis that allowed for the rejection of the null hypothesis and the confirmation that the effectiveness interval is statistically significant.

## Druggies
### Druggies Graphical Data
```{r include=FALSE,echo=FALSE}
Drugusers <- FinalData %>% filter(DrugUser=="yes")

Drugusers$treatment <- factor(Drugusers$treatment, levels = c("placebo", "vaccine"))
```

```{r echo=FALSE}
barchartGC(~infected + treatment,data=Drugusers)
barchartGC(~infected + treatment,data=Drugusers, type="percent")
```

In the Druggies data set there was a noticeable difference to the two data sets of the male and female genders; however the Druggies data set closely parallels the LGBTQ data set. The Druggies data set actually show an increase in Covid-19 infection rates in those who received the vaccine. However similarly to all the other data sets the total numerical difference in infection rates is very small despite being a large percentage difference.

### Druggies Numerical Data
```{r echo=FALSE}
table1 <- xtabs(~infected + treatment, data=Drugusers)
rowPerc(table1)
colPerc(table1)
```

The somewhat large increase in percent in infection rates within the Druggies sub-population between those who received the drug and those who received the placebo(3.1% of those given the actual vaccine were infected versus 0.4% of those who received the placebo) is a very poor initial indication that the vaccine effectively decreases overall infection rates of Covid-19 in the Druggies sub-population in concurrence with what was observed in the graphical data. It also closely follows with the similarities that were seen between the LGBTQ graphical data and Druggies.

### Druggies Inferential Data
```{r echo=FALSE}
chisq.test(table1)
fisher.test(table1)


fish <- fisher.test(table1)


low=(fish$conf.int[1]/(fish$conf.int[1] + 1))*100
hi=(fish$conf.int[2]/(fish$conf.int[2] + 1))*100
options(digits=2)

low
hi
```

The very low p-values in the chi-squared test and fisher exact test(3x10^-9 and 2x10^-10 respectively) indicate that the difference in infection rates is indeed statistically significant which means the drug is likely ineffective at reducing infection rates in the drug user population and allows us to reject the null hypothesis. The effectiveness interval for the vaccine in the drug user population of 4.6%-22% is far below the approximate 50% threshold that is desired for vaccine effectiveness.

### Druggies Conclusion and Discussion

The vaccine in the drug user sub-population appears to be completely ineffective to the point of potentially increasing infection rates; having a effectiveness interval of 4.6%-22% which indicates that it is an effective vaccine at preventing Covid-19 infections in the drug user sub-population. This is confirmed by the very small p values obtained in the inferential data analysis that allowed for the rejection of the null hypothesis and the confirmation that the effectiveness interval is statistically significant.

# Overall Results and Conclusions

The vaccine appears to be effective at preventing Covid-19 infections in the overall population though slightly less so in females than it is in males. However in the LGBTQ and drug user sub-populations the vaccine is highly ineffective to the point of potentially increasing infection rates of Covid-19 though it is very slightly more effective for the LGBTQ sub-population than it is for drug users. The null hypothesis was able to be rejected for all groups looked at in this investigation, though in a negative light for the LGBTQ and drug user sub-populations. Overall the vaccine is most effective in male subjects, followed by female subjects, followed by LGBTQ subjects, and finally it is least effective in drug users.